Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-03-01 paper.tex
Remove Macros
Add anonymous repo.
Final Edits.

We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may continuously spin without making progress, resulting in \emph{starvation}---consuming execution resources that would otherwise allow the blocking tile to complete its reduction and stalling the kernel indefinitely.

@@

LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, OBE guarantees forward progress.

@@

For a subgroup of size $s$, the Kogge-Stone network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s)                                                                \\
                & = O(n).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes. However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
\section{Discussion}

@@

\subsection{Limits on Speed-of-Light Performance}
Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling. Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by memory loads and stores. The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling. Another potential bottleneck is the speed of atomic operations. On GPUs where atomics do not operate within the cache, the resulting higher latency leads to delayed tile updates, which in turn increases lookback lengths.

@@

\subsection{Conclusion}
In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints. Our approach enables any workgroup to compute the reduction of any work tile, eliminating the risk of starvation. We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee. Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG. Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.

Our artifact is available in the following GitHub repository: TODO. The repository contains per API requirements and building instructions.

>>>>

We present \emph{Decoupled Fallback}, a fully portable \emph{Chained Scan} capable of achieving SOL performance without requiring forward-progress guarantees. Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's scheduling model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may continuously spin without making progress, resulting in \emph{starvation}---consuming execution resources that would otherwise allow the blocking tile to complete its reduction and stalling the kernel indefinitely.

@@

LOBE adds an additional condition on top of OBE\@: lower-ID workgroups must be scheduled before higher-ID ones. However, this condition is unnecessary for \emph{Chained Scan} for two reasons: (1) We already guarantee serial ordering of work items through atomic bumping, making OBE sufficient. (2) \emph{Chained Scan} only requires forward progress after a workgroup has atomically acquired its tile index. By definition, a workgroup that has acquired its index must be occupant on a physical processor (i.e., it has made at least one execution step); thus, the OBE guarantee is sufficient for forward progress.

@@

For a subgroup of size $s$, the Kogge-Stone subgroup network has a work complexity of $s \log_2 s$. Given an input of size $n$, the work complexity is:
\begin{equation}
  \begin{aligned}[b]
    \text{Work} & = \underbrace{\vphantom{\sum_{k=0}^{\lceil \log_s n \rceil}}n}_{\text{fanout}}
    +
    \underbrace{\sum_{k=1}^{\lceil \log_s n \rceil}}_{\text{iterations}}
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}\frac{n}{s^k}}_{\text{calls per iteration}}
    \cdot
    \underbrace{\vphantom{\sum_{k=1}^{\lceil \log_s n \rceil}}s \log_2 s}_{\text{work per call}} \\
                & = n + s \log_2 s \cdot \sum_{k=1}^{\lceil \log_s n \rceil} \frac{n}{s^k}       \\
                & = O\left(n + n \log_2 s\right)                                                 \\
                & = O(n \log_2 s)                                                                \\
                & = O(n).
  \end{aligned}
\end{equation}
Since the subgroup size remains constant within each kernel launch, the final work complexity is $O(n)$. The Kogge-Stone scan has a depth of $\log_2 s$; the total depth of the workgroup-wide scan is:
\begin{equation}
  \begin{aligned}[b]
    \text{Depth} & = \underbrace{\lceil \log_s n \rceil}_{\text{loop iterations}}
    \cdot \underbrace{\log_2 s}_{\text{depth per iteration}}                           \\
                 & = \left\lceil \frac{\log_2 n}{\log_2 s} \right\rceil \cdot \log_2 s \\
                 & = \log_2 n + O(1).
  \end{aligned}
\end{equation}
Thus, this network offers several advantages: minimal depth of $\log_2 n$; asymptotically optimal $O(n)$ size; and zero bank conflicts across all supported subgroup sizes. However, while minimal depth is preserved regardless of the subgroup size, this result can be somewhat misleading, as the depth incurred during the subgroup portion of the scan requires no barrier, whereas the main loop incurs a workgroup-wide barrier at each iteration. Consequently, while the theoretical depth remains constant across subgroup sizes, smaller subgroups tend to perform worse in practice due to the increased relative cost of synchronization and the overhead of additional iterations in the main loop.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups are scheduled for execution, reducing opportunities for context switching to mitigate latency.
\section{Discussion}

@@

\subsection{Limits on Speed-of-Light Performance}
Due to its increased latency, synchronization overhead, and longer critical path, \emph{Decoupled Fallback} exacerbates existing challenges faced by \emph{Decoupled Lookback}, particularly a lack of compute capacity and unfair scheduling. Devices that already lack sufficient compute resources to reach SOL performance are especially affected, as the additional compute overhead from fallback operations cannot be hidden by global loads and stores. The worst-case scenario arises in devices that suffer from both low compute power and unfair scheduling. Another potential bottleneck is the speed of atomic operations. On GPUs where atomics do not operate within the cache, the resulting higher latency leads to delayed tile updates, which in turn increases lookback lengths.

@@

\subsection{Conclusion}
In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints. Our approach enables any workgroup to compute the reduction of any blocking tile, eliminating the risk of starvation. We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee. Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG. Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.

Our artifact is available in the following GitHub repository: https://anonymous.4open.science/status/Decoupled-Fallback-Paper-A33C. The repository contains per API requirements and building instructions.
